% ************************** Thesis Acknowledgements **************************

\begin{acknowledgements}      

%Research contained in this thesis wouldn't have been possible without people around me. I would like to use this page to name a few, with many apologies to who 

%I wrote this thesis myself 


If this thesis was to be shared between everyone who made it happen, my share would be about a page. I would like to use this page to thank them and apologise to those who were not mentioned.

I would first like to thank my supervisors, James Fergusson and Paul Shellard. James has been the most patient and considerate supervisor one can ever hope for and was always there to give me valuable advice whenever I needed it. I am lucky to have been introduced to cosmology by him and benefit from his scientific intuition on many subjects. I am grateful to  Paul for his insightful comments throughout my studies and for running our department. I would like to thank Kacper Kornet for his technical support; he is the reason why my code could run at all. I also express my gratitude to Trinity College for creating such an inspiring environment to study and for providing financial support during my postgraduate studies.

I am immensely grateful to Philip Clarke for helping me throughout the ups and downs of my PhD as a close collaborator, office mate, badminton buddy and an awesome friend. DAMTP is the place I will miss forever thanks to Áron Kovács, Fred Alford, Owain Salter Fitz-Gibbon, Miren Radia, Roland Bittleston, Daniel Zhang, João Melo, Charles Board, Roxana and Mycroft Rosca-Mead, and many more amazing people who I have had the privilege to meet.

I am forever indebted to the members of the Donut who accompanied me throughout the pandemic. Chuyi Yang, Kaimyn Chapman, Paul Tao, and Zen Ng\textemdash thank you for making me feel at home during the wildest of times.

Last but not least, I am incredibly privileged to have a caring family. Mom, dad, and Hyojie, your bottomless love and boxes full of Korean food sent from five thousand miles away gave me the energy to keep moving forward. And grandpa, sorry I am a little late, but I hope you get to celebrate your grandson's PhD degree all the same from where you are.



\end{acknowledgements}
